## The Costs, Both Economic and Human, of Weather Events in the US
Here is a synopsis. Add several sentences.

### Data Processing
I got the data from here. Quick description. 

```{r, cache=TRUE}
stormdata <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
```

Let's look at the variable names.
```{r}
names(stormdata)
```

For this analysis, we care most about events and their human and economic costs across the United States. Let's take a look at those variables. 

First, the events.
```{r}
unique(stormdata$EVTYPE)[1:5]
```

Nearly 1000 different event types and those types are as expected. Should we aggregate similar types?

Next, the human costs. There are two variables that matter to us. 
```{r}
str(stormdata$FATALITIES)
summary(stormdata$FATALITIES)
str(stormdata$INJURIES)
summary(stormdata$INJURIES)
```

No surprises there. 

We will create a new variable called `CASUALTIES` which we define as the sum of `INJURIES` and 10 * `FATALITIES`. In estimating human cost, a single fatality should count much more than an injury.
```{r}
suppressMessages(library(dplyr))
#Start creating a subset of stormdata for specific analysis
stormdatatidy <- mutate(stormdata, CASUALTIES = FATALITIES * 10 + INJURIES)
summary(stormdatatidy$CASUALTIES)
```

Let's turn our attention to economic costs. Both property and crop damage count are economic costs.

First, property damage. Curiously, there are two variables which may matter.
```{r}
str(stormdata$PROPDMG)
summary(stormdata$PROPDMG)
str(stormdata$PROPDMGEXP)
summary(stormdata$PROPDMGEXP)
```

`PROPDMG` is a numeric vector as expected although the numbers seem low. Surprisingly, `PROPDMGEXP` is a 19 level factor variable. Let's take a closer look.
```{r}
summarize(group_by(stormdata, PROPDMGEXP), count = n())
```

Almost all observations are either `K`, `M` or `EMPTY`. It's reasonable to assume that `K` stands for thousands, `M` for millions and `EMPTY` for small (< $1000) amounts of damage or none at all. Let's take a closer look. 
```{r}
summarize(group_by(stormdata, PROPDMGEXP), count = format(n(), big.mark = ","), dollars = format(sum(PROPDMG), big.mark = ",", digits = "0", scientific = FALSE))
```

Given how large the `K`, `M` and `EMPTY` are in terms of dollars, we can safely ignore the remaining 16 factors as transcription errors. 

In our `stormdatatidy` set, let's create a new variable based on `PROPDMG` so that we have the damage in consistent dollars. Then, we'll keep only the observations we the three factors we care about.
```{r}
#Chaining would be nice to use here if I had the time.
stormdatatidy <- mutate(stormdatatidy, MULT = ifelse(PROPDMGEXP == "", 1, ifelse(PROPDMGEXP == "K", 1000, ifelse(PROPDMGEXP == "M", 1000000, 0))))
stormdatatidy <- mutate(stormdatatidy, propdamage = MULT * PROPDMG)
stormdatatidy <- filter(stormdatatidy, PROPDMGEXP == "" | PROPDMGEXP == "K" | PROPDMGEXP == "M")
dim(stormdata)
dim(stormdatatidy)
```

Now, we turn our attention to crop damage. Like property damage, there are two variables involved: `CROPDMG` and `CROPDMGEXP`. Let's use the same approach. 
```{r}
summarize(group_by(stormdata, CROPDMGEXP), count = format(n(), big.mark = ","), dollars = format(sum(CROPDMG), big.mark = ",", digits = "0", scientific = FALSE))
```

We see `K`, `M` and `EMPTY` dominate in the case, too. Therefore we will use the same approach as with property damage: create a new varible with consistent dollars and discard the presumed transcription errors. 
```{r}
stormdatatidy <- stormdatatidy %>%
        mutate(MULT = ifelse(CROPDMGEXP == "", 1, 
            ifelse(CROPDMGEXP == "K", 1000, 
            ifelse(CROPDMGEXP == "M", 1000000, 0)))) %>%
        mutate(cropdamage = MULT * PROPDMG) %>%
        filter(CROPDMGEXP == "" | CROPDMGEXP == "K" | CROPDMGEXP == "M")
dim(stormdata)
dim(stormdatatidy)
```

Now we'll sum `cropdamage` and `propdamage` to create overall `econdamage`.
```{r}
stormdatatidy <- stormdatatidy %>%
        mutate(econdamage = cropdamage + propdamage)
summary(stormdatatidy$econdamage)
```

Now, a final cleanup to capture just the variables we want.
```{r}
stormdatatidy <- select(stormdatatidy, EVTYPE, casualties = CASUALTIES, econdamage)
dim(stormdata)
dim(stormdatatidy)
```

Looks like we're ready to do some analysis. (Finally.)
```{r}
summary(stormdatatidy)
```
### Results
We will create with a simple summary table as the backbone of our analysis.
```{r}
sumtbl <- summarize(group_by(stormdatatidy, EVTYPE), casualties = sum(casualties), damage = sum(econdamage))
```

Let's see which events create the most casualties.
